---
title: "Demographic and Socioeconomic Predictors of Algebra 1 End-of-Course Achievement in Florida School Districts"
author: "Wendy Matende, Madison Edwards, Jegg Ononiewu"
bibliography: references.bib
format: 
    revealjs:
        embed-resources: true
        editor: visual
        theme: dracula
        fontsize: "275%"
        incremental: true
        transition: fade
        slide-number: true
        chalkboard: false
        width: 1920
        height: 1080
        auto-stretch: false
        execute: 
          echo: false
          warning: false
          message: false
---

```{r}
#| include: false
library(dplyr)
library(ggplot2)
library(tidyr)
library(quantreg)
library(corrplot)

final_data2 <- readRDS("final_data2.rds")
final_data2$Mean_Score <- as.numeric(trimws(final_data2$Mean_Score))
final_data2$Level3Plus <- as.numeric(final_data2$Level3Plus)
```

## The Gateway: Algebra 1 EOC

- **The Gateway Course:** Algebra 1 shapes student access to advanced mathematics, college readiness, and future STEM pathways.
- **High Stakes in Florida:**
- Must pass the **End-of-Course (EOC)** assessment to earn a standard high school diploma.
- Accounts for **30%** of a student's final course grade.
- **Systemic Impact:** Directly influences school accountability, remediation funding, and instructional support.

## District-Level Disparities

- Florida’s 67 school districts vary dramatically in demographic and socioeconomic composition.
- **Key Contextual Predictors:**
- **Demographics:** Racial/ethnic distributions (White, Black, Hispanic).
- **Socioeconomics:** Poverty, unemployment, median household income, and bachelor's degree attainment.
- **The Equity Problem:** Resource-dense districts provide structural advantages, while economically disadvantaged communities face systematic barriers.

## The Methodological Gap: Beyond the Mean

> "One-size-fits-all statistical models can mask meaningful differences among lower-, middle-, and higher-performing groups." — Reeves & Lowe (2009)

- **The Problem with OLS:** Traditional mean-based regression assumes that demographic and socioeconomic impacts are *uniform* across all districts.
- **The Reality:** Contextual influences vary. Factors affecting lower-achieving districts may operate differently in high-achieving ones (Tian, 2006; Haile & Nguyen, 2008).
- **The Florida Gap:** Existing state studies (e.g., Holley, 2017) rely heavily on mean-based methods, leaving a critical gap in distribution-sensitive analysis.

## The Solution: Quantile Regression

::::: columns
::: {.column width="50%"}
### OLS Regression

- Estimates the **conditional mean**.
- Assumes constant predictor effects.
- Can obscure variation in the tails of the distribution.
:::

::: {.column width="50%"}
### Quantile Regression

- Estimates relationships at **multiple points** ($\tau = 0.25, 0.50, 0.75$).
- Uncovers shifting magnitudes or changing signs across performance tiers.
:::
:::::

## Research Questions

1.  How are district-level **demographic characteristics** associated with Algebra 1 EOC achievement?
2.  How are district-level **socioeconomic characteristics** associated with Algebra 1 EOC achievement?
3.  Do the direction, magnitude, and significance of these predictors **vary across the achievement distribution**?
4.  Which predictors demonstrate the strongest associations at **specific quantiles**, and what are the implications for equity-focused planning?

## Study Significance & Contributions

- **For Policymakers:** Shifts focus from statewide averages to targeted resource allocation for struggling vs. excelling performance tiers.
- **For Educators:** Deepens the understanding of how community-level structural barriers uniquely pressure mathematics achievement.
- **For Researchers:** Fills a documented literature gap by applying a distribution-sensitive lens to Florida's K–12 educational outcomes.

## Quantile Regression

**Why Quantile Regression?**

* Developed by Koenker and Bassett (1978).
* Estimates conditional quantiles rather than only the conditional mean.
* Shows how predictor effects vary across different parts of the outcome distribution.
* Provides a more complete understanding of relationships than Ordinary Least Squares (OLS).

**Key References**

* Koenker & Bassett (1978)
* Reeves & Lowe (2009)
* Costanzo & Desimoni (2017)

---

## Quantile Regression Model

**General Quantile Regression Model**

$$Q_Y(\tau|X)=X^{T}\beta(\tau)$$

**where**

* $Q_Y(\tau|X)$ = conditional quantile of the response variable
* $X$ = explanatory variables
* $\beta(\tau)$ = regression coefficients for quantile $\tau$

**Quantiles Estimated**

* 10th percentile ($\tau = 0.10$)
* 25th percentile ($\tau = 0.25$)
* 50th percentile ($\tau = 0.50$)
* 75th percentile ($\tau = 0.75$)
* 90th percentile ($\tau = 0.90$)

---

## Estimation Method

**Quantile Regression Estimation**

The regression coefficients are estimated by minimizing the check-loss function:

$$\hat{\beta}(\tau) = \arg\min_{\beta} \sum_{i=1}^{n} \rho_{\tau}(y_i-x_i^{T}\beta)$$

**where**

$$\rho_{\tau}(u)=u[\tau-I(u<0)]$$

**Key Features**

* Minimizes weighted absolute residuals.
* Uses linear programming instead of least squares.
* Robust to outliers.
* Does not require normality or constant error variance.
* Models relationships across the entire outcome distribution.

---

## Statistical Implementation

**Statistical Analysis**

* Quantile regression models are estimated using the `quantreg` package in R.
* Linear programming algorithms are used to estimate regression coefficients.
* Bootstrap standard errors are used for statistical inference.
* Results are compared with Ordinary Least Squares (OLS) regression.

------------------------------------------------------------------------

## Results and Discussion

### Purpose of the Analysis

This section presents the findings in the order the analyses were conducted.

::: incremental
- Descriptive analysis
- Relationships between socioeconomic variables and achievement
- OLS regression
- Quantile regression
- Policy implications
:::

## Distribution of Algebra 1 Mean Scores

:::::::: columns
::: {.column width="45%"}
### Key Findings & Interpretation

- **Clustering:** Mean scores clustered tightly between \~390 and 405.
- **Low Variation:** District variation across Florida was relatively small.
- **Meaningful Gaps:** Despite a compressed range, small variations reflect key socioeconomic inequalities (@reardon2019geography, @sirin2005socioeconomic).

**Answer to Research Question 1:**

✔ Baseline district achievement varies within a narrow but meaningful distribution.
:::

:::::: {.column width="55%"}
```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
#| warning: false
ggplot(final_data2, aes(x = Mean_Score)) +
geom_histogram(bins = 10, fill = "steelblue", color = "white") +
labs(
title = "Distribution of Algebra 1 Mean Scale Scores",
x = "Mean Scale Score",
y = "Frequency"
) +
theme_minimal(base_size = 24)

```
:::

::::::::


## Poverty and Algebra 1 Achievement

::::: columns
::: {.column width="45%"}
### Key Findings & Interpretation

- **Negative Relationship:** Higher-poverty districts generally exhibited lower mean scores.
- **Confounded Effect:** Overlaps with income and education; loses significance in multivariate OLS models.
- **Structural Variable:** Reflects broad structural disadvantage rather than operating independently (@sirin2005socioeconomic).

**Answer to Research Question 2:**

✔ Poverty is negatively associated with achievement, but lacks an independent multivariate effect.
:::
::: {.column width="55%"}
```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
#| warning: false
ggplot(final_data2, aes(x = poverty_rate, y = Mean_Score)) +
geom_point(size = 3, alpha = 0.7, color = "darkred") +
geom_smooth(method = "lm", color = "black") +
labs(
title = "Poverty and Algebra 1 Achievement",
x = "Poverty Rate",
y = "Mean Scale Score"
) +
theme_minimal(base_size = 24)

```
:::

::::::::

## Median Household Income

::: {.columns}
::: {.column width="45%"}
### Key Findings & Interpretation
* **Positive Bivariate Trend:** Higher-income districts show structurally better performance trends (@duncan2016rising).
* **Multicollinerality:** Insignificant in OLS due to a high correlation with community educational attainment levels.

**Answer to RQ 2:**

✔ Median household income shows a positive bivariate relationship but not an independent multivariate effect.
:::

::: {.column width="55%"}
```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
#| warning: false
ggplot(final_data2, aes(x = median_income, y = Mean_Score)) +
geom_point(size = 3, alpha = 0.7, color = "darkgreen") +
geom_smooth(method = "lm", color = "black") +
labs(
title = "Median Income and Algebra 1 Achievement",
x = "Median Household Income",
y = "Mean Scale Score"
) +
theme_minimal(base_size = 24)

```
:::

::::::::

## Bachelor’s Degree Attainment

:::::::: columns
::: {.column width="45%"}
### Key Findings & Interpretation
- **Strong Predictor:** Robust positive association across the entire distribution.
- **Strongest Impact:** Effect intensifies at higher-performing districts.
- **Literature Match:** Confirms parental and community educational capital drives student success (@reardon2019geography).

**Answer to Research Question 2:**

✔ Bachelor's degree attainment was one of the strongest predictors of achievement.
:::

::: {.column width="55%"}
```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
#| warning: false
ggplot(final_data2, aes(x = bachelors_rate, y = Mean_Score)) +
geom_point(size = 3, alpha = 0.7, color = "purple") +
geom_smooth(method = "lm", color = "black") +
labs(
title = "Bachelor’s Degree Attainment and Achievement",
x = "Bachelor’s Degree Rate",
y = "Mean Scale Score"
) +
theme_minimal(base_size = 24)

```
:::

::::::::

## Unemployment and Achievement

:::::::: columns
::: {.column width="45%"}
### Key Findings & Interpretation
- **Consistent Harm:** Higher unemployment tracks lower achievement.
- **Remains Significant:** Holds strong and negative in multivariate OLS and across all quantiles.
- **Context:** Economic instability directly impacts localized student learning environments (@duncan2016rising).
**Answer to Research Question 2:**

✔ Unemployment was a strong and consistent negative predictor.
:::

::: {.column width="55%"}
```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
#| warning: false
ggplot(final_data2, aes(x = unemployment_rate, y = Mean_Score)) +
geom_point(size = 3, alpha = 0.7, color = "orange") +
geom_smooth(method = "lm", color = "black") +
labs(
title = "Unemployment and Algebra 1 Achievement",
x = "Unemployment Rate",
y = "Mean Scale Score"
) +
theme_minimal(base_size = 24)
```
:::
::::::::

## Ordinary Least Squares (OLS) Results

:::::::: columns
::: {.column width="50%"}
### Predictor Breakdown
* **Significant Predictors:**
* 👍 **Bachelor’s Degree Attainment** (Positive impact)
* 👎 **Unemployment Rate** (Negative impact)

* **Non-Significant Predictors:**
* Poverty, Median Income, % White, % Black, and % Hispanic (primarily due to overlapping explanatory variances).
:::

::: {.column width="50%"}
### Interpretation & Limitations
* **Average Relationships:** OLS evaluates global, average effects across your entire dataset exclusively.
* **Uniformity Bias:** It assumes identical effects across all districts, masking crucial differences at the top and bottom tails (@koenker1978).

**Answer to Research Question 3:**
✔ OLS successfully described average trends, but completely missed unique shifts across different school achievement levels.
:::
::::::::
```{r}
#| echo: false
#| warning: false
#| results: 'hide'
# Running the model behind the scenes so the data is ready for upcoming slides
ols_model2 <- lm(
Mean_Score ~ pct_white + pct_black + pct_hispanic +
poverty_rate + unemployment_rate + median_income + bachelors_rate,
data = final_data2
)

```

## Why Quantile Regression?

:::::::: columns
::: {.column width="50%"}
### Beyond the Mean
* **OLS Limitation:** OLS only estimates relationships at the conditional mean.
* **Quantile Advantage:** Quantile regression models relationships at specific, distinct points across the entire distribution.

### Achievement Levels Modeled ($\tau$)
* **$\tau = 0.10$ & $0.25$:** Low-Achieving Districts
* **$\tau = 0.50$:** Median-Achieving Districts
* **$\tau = 0.75$ & $0.90$:** High-Achieving Districts
:::

::: {.column width="50%"}
### Analytical Advantages

1. **Context-Dependent Dynamics:** Identifies if predictors vary based on whether a district is low-, middle-, or high-performing.
2. **Unmasks Hidden Patterns:** Reveals critical structural patterns that are hidden completely by standard OLS averages.
:::
::::::::

## Lowest-Achieving Districts ($\tau = 0.10$)

```{r}
#| echo: false
#| warning: false
#| message: false
# Running quantile regression models in the background
library(quantreg)
rq10 <- rq(Mean_Score ~ pct_white + pct_black + pct_hispanic + poverty_rate + unemployment_rate + median_income + bachelors_rate, tau = .10, data = final_data2)
rq25 <- rq(Mean_Score ~ pct_white + pct_black + pct_hispanic + poverty_rate + unemployment_rate + median_income + bachelors_rate, tau = .25, data = final_data2)
rq50 <- rq(Mean_Score ~ pct_white + pct_black + pct_hispanic + poverty_rate + unemployment_rate + median_income + bachelors_rate, tau = .50, data = final_data2)
rq75 <- rq(Mean_Score ~ pct_white + pct_black + pct_hispanic + poverty_rate + unemployment_rate + median_income + bachelors_rate, tau = .75, data = final_data2)
rq90 <- rq(Mean_Score ~ pct_white + pct_black + pct_hispanic + poverty_rate + unemployment_rate + median_income + bachelors_rate, tau = .90, data = final_data2)
```

:::::::: columns

::: {.column width="50%"}

### Main Findings

* **Demographic Sensitivity:** Race-related variables showed their strongest negative effects at this tail:
* `pct_black`: -0.88331
* `pct_white`: -0.83163
* `pct_hispanic`: -0.81958

* **Economic Instability:** Unemployment acts as a highly severe negative predictor at this specific level.

:::

::: {.column width="50%"}

### Interpretation & Insights

* **Structural Disadvantage:** Low-performing school districts experience the most intense compounding influence from systemic disadvantages.
* **Compounded Inequalities:** Racial disparities at this level are heavily connected with structural and environmental inequality factors (@owens2017income).

#### Answer to Research Questions 3 & 4:

* **Key Takeaway:** Predictor effects are strongest among the lowest-performing districts; demographic makeup and localized unemployment heavily dominate performance outcomes at the bottom tail.

:::

::::::::


## Middle Achievement Districts ($\tau = 0.25$ & $\tau = 0.50$)

:::::::: columns
::: {.column width="50%"}
### Main Findings

::: {.incremental}
- **Demographic Fade:** Race-related variable coefficients weaken substantially compared to the lowest tail.
- **Rise of Education Capital:** The impact of Bachelor's degree attainment rates increases.
- **Persistent Stress:** Unemployment continues to exert a sturdy negative pull on results.
:::
:::

::: {.column width="50%"}
### Interpretation & Insights

* **Shifting Dynamics:** As districts pull closer toward state averages, structural-demographic vulnerabilities diminish slightly.
* **Community Capital:** Local educational assets emerge as stronger, more dominant engines of student performance (@reardon2019geography).

**Answer to Research Questions 3 & 4:**
✔ Predictor effects change across the distribution, with educational capital becoming increasingly influential in middle-performing systems.
:::
::::::::

## Highest Achievement Districts ($\tau = 0.75$ & $\tau = 0.90$)

:::::::: columns
::: {.column width="50%"}
### Main Findings

* **Dominant Asset:** Bachelor’s degree attainment reaches its peak effect size (**0.47533** at **$\tau = 0.90$**).
* **Demographics Neutered:** Race and demographic composition become completely non-significant at this tail.
* **Constant Harm:** Unemployment remains a steady, unyielding negative factor even in these top-tier systems.
:::

::: {.column width="50%"}
### Interpretation & Insights

* **Educational Concentration:** High-performing districts are primarily distinguished and sustained by their localized concentration of adult educational capital.
* **Advantage Accumulation:** This pattern strongly reinforces an "accumulation of advantage" framework across top-performing student cohorts (@bryk2015accelerating).

**Answer to Research Questions 3 & 4:**
✔ Educational attainment completely dominates among high-achieving districts, while demographic composition drops entirely in predictive power.
:::
::::::::

## Correlation Matrix

:::::::: columns
::: {.column width="40%"}
### Key Findings & Core Logic

* **Positives:** Test scores track heavily with Income (**r = 0.72**) and Bachelor's degrees (**r = 0.65**).
* **Negatives:** Poverty and Income share a severe inverse relationship (**r = -0.82**).
* **Collinearity:** Income vs. Bachelor's degree rates are profoundly linked (**r = 0.82**).

### Interpretation & Insights

* **Systemic Interconnection:** Socioeconomic predictors are deeply interconnected, rather than operating in isolation (@sirin2005socioeconomic).
* **OLS Dropouts:** This severe multicollinearity directly explains why powerful individual variables drop out of statistical significance when thrown together into a multi-variable OLS model.

**Answer to Research Questions 1 & 2:**
✔ District achievement scores are molded by clusters of interconnected context, not single isolated variables.
:::

::: {.column width="60%"}

```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(corrplot)

corr_data <- final_data2 %>%
select(Mean_Score, pct_white, pct_black, pct_hispanic,
poverty_rate, unemployment_rate, median_income, bachelors_rate)

cor_matrix <- cor(corr_data, use = "pairwise.complete.obs")

corrplot(
cor_matrix, method = "color", type = "upper",
tl.cex = 0.8, tl.col = "white", number.cex = 0.7,
addCoef.col = "black",
col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
mar = c(0,0,1,0)
)
```

:::
::::::::

## Quantile Regression Coefficient Plot

:::::::: columns
::: {.column width="40%"}
### Main Dynamic Patterns

1. **Unemployment:** Consistently flat and negative across every single quantile layer.
2. **Bachelor's Degrees:** Increases steadily, escalating from lowest to highest quantiles.
3. **Demographics:** Exhibits its strongest drag at low quantiles, then levels off at higher quantiles.

### Interpretation & Insights

* **Visual Proof:** This visual trend confirms the absolute existence of strong heterogeneous effects across the achievement distribution, which traditional mean-based OLS completely hides (@koenker1978).

**Answer to Research Question 3:**
✔ Quantile regression successfully revealed heterogeneous effects that were completely ignored or missed by standard OLS.
:::

::: {.column width="60%"}
```{r}
#| fig-width: 12
#| fig-height: 7
#| echo: false
#| warning: false
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)

coef_data <- data.frame(
Quantile = c(.10, .25, .50, .75, .90),
pct_white = c(coef(rq10)["pct_white"], coef(rq25)["pct_white"], coef(rq50)["pct_white"], coef(rq75)["pct_white"], coef(rq90)["pct_white"]),
pct_black = c(coef(rq10)["pct_black"], coef(rq25)["pct_black"], coef(rq50)["pct_black"], coef(rq75)["pct_black"], coef(rq90)["pct_black"]),
pct_hispanic = c(coef(rq10)["pct_hispanic"], coef(rq25)["pct_hispanic"], coef(rq50)["pct_hispanic"], coef(rq75)["pct_hispanic"], coef(rq90)["pct_hispanic"]),
poverty_rate = c(coef(rq10)["poverty_rate"], coef(rq25)["poverty_rate"], coef(rq50)["poverty_rate"], coef(rq75)["poverty_rate"], coef(rq90)["poverty_rate"]),
unemployment_rate = c(coef(rq10)["unemployment_rate"], coef(rq25)["unemployment_rate"], coef(rq50)["unemployment_rate"], coef(rq75)["unemployment_rate"], coef(rq90)["unemployment_rate"]),
median_income = c(coef(rq10)["median_income"], coef(rq25)["median_income"], coef(rq50)["median_income"], coef(rq75)["median_income"], coef(rq90)["median_income"]),
bachelors_rate = c(coef(rq10)["bachelors_rate"], coef(rq25)["bachelors_rate"], coef(rq50)["bachelors_rate"], coef(rq75)["bachelors_rate"], coef(rq90)["bachelors_rate"])
)

coef_long <- coef_data %>%
pivot_longer(cols = -Quantile, names_to = "Variable", values_to = "Coefficient")

ggplot(coef_long, aes(x = Quantile, y = Coefficient, color = Variable)) +
geom_line(linewidth = 1.2) +
geom_point(size = 3) +
labs(
title = "Quantile Regression Estimates Across Quantiles",
x = "Quantile Level (\u03c4)",
y = "Coefficient Estimate"
) +
theme_minimal(base_size = 24) +
theme(legend.position = "bottom")
```

:::

::::::::

## Overall Quantile Regression Findings

:::::::: columns
::: {.column width="50%"}
### Summary of Distributional Realities

::: {.incremental}
- **At the Bottom ($\tau = 0.10$):** Demographic variables and systemic marginalization matter most to performance outcomes.
- **At the Top ($\tau = 0.90$):** Regional adult educational capital dominates and drives exceptional achievement.
- **Across the Entire Board:** Localized unemployment acts as a consistent, unyielding negative barrier to student success.
:::
:::

::: {.column width="50%"}
### Core Presentation Takeaway

* **Beyond Global Averages:** Educational inequality cannot be treated as a single, uniform number.
* **Multi-Layered Reality:** Structural barriers change their behavior depending on whether a school system is fighting to catch up or working to maintain its advantage.

**Research Questions Takeaway:**
✔ Research Questions 3 and 4 were answered comprehensively by demonstrating that predictor effects vary sharply across the achievement distribution. Educational inequality is structural, non-uniform, and deeply multi-layered.
:::
::::::::

## Policy Implications

| Finding | Policy Implication |
|:---|:---|
| **High Unemployment** | Expand tutoring, mental health, and attendance support |
| **Low Educational Attainment** | Invest in community education and workforce development |
| **Low-Achieving Districts** | Provide targeted interventions and wraparound services |
| **Different Effects Across Quantiles** | Avoid one-size-fits-all educational policies. School interventions must adjust based on performance distribution layers (@bryk2015accelerating) |

:::
::::::::

## Contribution, Limitations, and Future Research

:::::::: columns
::: {.column width="33%"}
### Contributions
- Applied quantile regression to Florida Algebra 1 performance data.
- Discovered heterogeneous effects missed by classic OLS.
- Advanced contextual educational inequality literature.
:::

::: {.column width="33%"}
### Limitations
- Cross-sectional data (no causal pathways).
- District-level aggregation (susceptible to ecological fallacy).
- Lacks school-level variables (funding details, teacher experience).
:::

::: {.column width="34%"}
### Future Work
- Longitudinal tracking.
- Multilevel modeling structures (students nested inside districts).
- Cross-state comparisons to test model generalizability.
:::
::::::::

## Conclusion

:::::::: columns
::: {.column width="100%"}
### Core Presentation Takeaways

* **Socioeconomic Primacy:** Localized socioeconomic profiles heavily dictate Algebra 1 outcomes across Florida.
* **The Stability Tax:** Unemployment consistently acts as an achievement depressor across all distribution percentiles.
* **Methodological Value:** Quantile regression surfaces critical, hidden policy insights that mean-based OLS models completely wipe out.

:::
::::::::


## 🎯 Final Take-Home Message

:::::::: columns
::: {.column width="100%"}
> **The Core Reality:** Educational inequality is not uniform across Florida school districts.

* **Distributional Insights:** Quantile regression provides a far more comprehensive and nuanced understanding of how socioeconomic and demographic factors influence achievement.
* **Tailored Policy:** By looking beyond the average, we can see exactly how these factors change their behavior at different student performance levels—allowing for more precise, targeted policy interventions.

:::
::::::::

## References

::: {#refs}
:::
::::::::


